---
title: "PA1_template.Rmd"
author: "Jesús Prada Alonso"
date: "24 de enero de 2016"
output: html_document
---


## Loading and preprocessing the data

### 1) Load the data

```{r}
data<-read.csv("D://R/input/coursera/activity.csv");
```

### 2) Process/transform the data (if necessary) into a format suitable for your analysis

This step was not necessary.

## What is mean total number of steps taken per day?

### 1) Calculate the total number of steps taken per day
We use tapply for this step. option na.rm is set to TRUE because data$steps contains
NA values.

```{r}
sums_by_day<-tapply(data$steps,data$date,sum,na.rm=TRUE);
sums_by_day[is.na(sums_by_day)]<-0;
```

### 2) Make a histogram of the total number of steps taken each day

```{r}
hist(sums_by_day);
```

### 3) Calculate and report the mean and median of the total number of steps taken per day
The means is computed by the following instructions:

```{r}
means_by_day<-tapply(data$steps,data$date,mean,na.rm=TRUE);
means_by_day[is.na(means_by_day)]<-0;
print(means_by_day);
```

The median is computed by the following instructions:

```{r}
medians_by_day<-tapply(data$steps,data$date,median,na.rm=TRUE);
medians_by_day[is.na(medians_by_day)]<-0;
print(medians_by_day);
```

## What is the average daily activity pattern?

### 1) Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

First, we compute the average number of steps taken each 5-minute interval.

```{r}
means_by_int<-tapply(data$steps,data$interval,mean,na.rm=TRUE);
means_by_int[is.na(means_by_int)]<-0;
```

Then we plot these values as a time series.

```{r}
plot(means_by_int ~ unique(data$interval),type="l",
     xlab="interval",ylab="average_steps")
```

### 2) Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

The 5-minute interval with the maximum averaged number of steps is

```{r}
names(which.max(means_by_int));
```

In this interval, the mean value is

```{r}
print(means_by_int[which.max(means_by_int)]);
```

## Imputing missing values

### 1) Calculate and report the total number of missing values in the dataset

```{r}
nrow(data) - sum(complete.cases(data))
```

### 2) Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

We opt to fill missing values using the mean for the corresponding 5-minute interval, as shown in the next step.

### 3) Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
new_data<-data;
new_data$steps[is.na(data$steps)]<-means_by_int[as.character(data$interval[is.na(data$steps)])]
```

### 4) Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

The total number of steps taken each day is

```{r}
new_sums_by_day<-tapply(new_data$steps,new_data$date,sum,na.rm=TRUE);
new_sums_by_day[is.na(new_sums_by_day)]<-0;
```

The histogram for these values is

```{r}
hist(new_sums_by_day);
```

the mean of steps taken per day is computed by the following instructions:

```{r}
new_means_by_day<-tapply(new_data$steps,new_data$date,mean,na.rm=TRUE);
new_means_by_day[is.na(new_means_by_day)]<-0;
print(new_means_by_day);
```



the median of steps taken per day is computed by the following instructions:

```{r}
new_medians_by_day<-tapply(new_data$steps,new_data$date,median,na.rm=TRUE);
new_medians_by_day[is.na(new_medians_by_day)]<-0;
print(new_medians_by_day);
```

Mean and median values change when imputing missing data. This can be clearly seen in their values for the first value of the dataset. If we print the absolute difference between the values with and without imputing data for this we obtain.

```{r}
print(sprintf("Mean diff: %f",abs(new_means_by_day[1]-means_by_day[1])));
print(sprintf("Median diff: %f",abs(new_medians_by_day[1]-medians_by_day[1])));
```

The total daily number of steps. This can also be seen in its value for the first value of the dataset


```{r}
print(sprintf("Sum diff: %f",abs(new_sums_by_day[1]-sums_by_day[1])));
```

And also in their corresponding histograms

```{r}
par(mfrow=c(1,2));
hist(sums_by_day);
hist(new_sums_by_day);
```

## Are there differences in activity patterns between weekdays and weekends?

### 1) Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

First, we create a vector containing the day of the week for all the dates in the dataset.

```{r}
dates<-as.Date(new_data$date);
day<-weekdays(dates);
```
Then we create a factor vector using this information.

```{r}
day_factor<-rep("weekday",length(day));
day_factor[((day == "sábado") | (day == "domingo"))]="weekend";
day_factor<-factor(as.character(day_factor),levels=c("weekday","weekend"));
```

### 2) Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

First we compute the average number of steps for weekday days.

```{r}
data_weekday<-new_data[day_factor=="weekday",];
new_means_by_int_1<-tapply(data$steps,data$interval,mean,na.rm=TRUE);
new_means_by_int_1[is.na(new_means_by_int_1)]<-0;
```

Then we compute this value for weekend days.

```{r}
data_weekend<-new_data[day_factor=="weekend",];
new_means_by_int_2<-tapply(data$steps,data$interval,mean,na.rm=TRUE);
new_means_by_int_2[is.na(new_means_by_int_2)]<-0;
```

Finally we compute a panel plot with two rows each one of them containing one figure using these values.

```{r}
par(mfrow=c(2,1));
plot(new_means_by_int_1 ~ unique(new_data$interval),type="l",
     xlab="interval",ylab="average_steps");
plot(new_means_by_int_2 ~ unique(new_data$interval),type="l",
     xlab="interval",ylab="average_steps");
```